scRNAseq benchmark in want of a title

### 2019-12-02 bmc_article.tex
Updates from Overleaf

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets). In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF]. 
Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \textit{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \textit{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs]. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested DoubletFinder [REF] and scran's doubletCells [REF], both of which use similarity to artificial doublets, and scds [REF], which relies on a combination of coexpression and binary classification. DoubletFinder integrates a thresholding based on the proportion of expected doublets, while scran and scds return scores which must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e. doublets from different cell types) more efficiently. We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B). Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \label{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \label{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in `scater` (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard Seurat clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (sctransform's variance-stabilizing transformation involves something analogous to scaling). 
Seurat's scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation with technical covariates.  sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling. The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.

@@

Finally, although some normalization methods had a tendency to lead to a higher (e.g. sctransform) or lower (e.g. stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).

@@

A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq. Seurat's earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types. Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance. We therefore compared these methods.

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations. Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard Seurat normalization (left) or sctransform (right). The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

@@

We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, scran's denoise PCA, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization. Seurat's default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.

@@

We next investigated the impact of the number of dimensions used on the separability of the subpopulations. Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact. We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA). As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition. We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields. Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method. Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization. Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e. maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations.

@@

@@

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: Seurat (version 3.0.0), sctransform (0.2.0), DoubletFinder (2.0.1), scds (1.0.0), pipComp (0.2). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

\subsection*{Simulated datasets}
The `simMix1` dataset was based on the human PBMC CITE-seq data deposited under `GSE100866`. Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat. We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant such subpopulations. We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations. The `simMix2` dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion. The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.

@@

\subsection*{Doublet detection method}
Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}. Version 0.1 was used here. Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

@@

The deviance explained by subpopulations was calculated directly on counts using the \textit{pipComp::getDevianceExplained} function. The function uses edgeR to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model. For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model. In the rare cases where this resulted in a negative deviance explained, it was set to zero.

>>>>

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF]. 
Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \texttt{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs]. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of coexpression and binary classification. \texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores which must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e. doublets from different cell types) more efficiently. We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B). Our method (\texttt{scDblFinder}) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \label{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \label{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \texttt{scater} (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \texttt{Seurat}, we tested \texttt{scran}'s pooling-based normalization \citep{lunPooling2016}, \texttt{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling). 
\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates.  \texttt{sctransform} had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling. The only exception is \texttt{Seurat} normalization with scaling regressing out the number of features, which surprisingly increased the correlation.

@@

Finally, although some normalization methods had a tendency to lead to a higher (e.g. \texttt{sctransform}) or lower (e.g. stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).

@@

A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq. \texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types. Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering the genes by transformed variance. We therefore compared these methods.

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations. Estimates of variance explained by subpopulations based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \texttt{Seurat} normalization (left) or \texttt{sctransform} (right). The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

@@

We next investigated the step of dimensionality reduction, comparing in particular \texttt{Seurat}'s PCA, \texttt{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization. \texttt{Seurat}'s default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.

@@

We next investigated the impact of the number of dimensions used on the separability of the subpopulations. Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact. We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA). As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition. We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields. Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method. Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization. Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e. maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations, suggesting that these components help classification despite being individually uninformative.

@@

We finally moved to the clustering step. Given previous work on the topic\citep{duoClustering2018} and the success of graph-based clustering methods for scRNA-seq, we simply compared Seurat to two \texttt{scran} SNN-based clustering approaches respectively using random walks (`walktrap') or optimizing the modularity score (`fast greedy'), with standard normalization or after \texttt{sctransform}.

(Interpretation of the results... walktrap top?)

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
    \caption{\textbf{Evaluation of clustering methods.}}
    \label{fig:figure9}
\end{figure}

\subsection*{Further extensions to the pipeline: imputation/denoising}

The basic pipeline used here can be extended by adding steps and using the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques on their impact on classification. Observing that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. While all methods improved the separability of some subpopulations, no method had a positive impact on the silhouette width of all populations (Figure \ref{fig:figure10}A). Similarly, not all methods performed as well across all datasets in terms of clustering accuracy (Figure \ref{fig:figure10}A). As expected, 10X datasets which have more cells at a lower coverage benefited more from these methods, which were instead rather deleterious in smart-seq datasets.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
    \caption{\textbf{Imputation/denoising methods. A:} Average silhouette width per subpopulation across various denoising/imputation methods. \textbf{B:} Clustering accuracy across denoising/imputation methods.}
    \label{fig:figure10}
\end{figure}

@@

\subsection*{Concrete recommendations}

\subsection*{Limitations and open questions}

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

\subsection*{Simulated datasets}
The `simMix1' dataset was based on the human PBMC CITE-seq data deposited under `GSE100866'. Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat. We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant such subpopulations. We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations. The `simMix2' dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion. The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.

@@

\subsection*{Doublet detection method}
Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}. Version 0.1 was used here. Briefly, we first cluster the cells using \texttt{scran}'s \texttt{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artificial doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

@@

The deviance explained by subpopulations was calculated directly on counts using the \texttt{pipeComp::getDevianceExplained} function. The function uses edgeR to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model. For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model. In the rare cases where this resulted in a negative deviance explained, it was set to zero.
